---
title: "Worker Tool for Computational Studies"
date: 2023-02-12
categories: programming
---

In a large statistical simulation or other computational study, it can be
labor intensive to kick off all of the runs whose results will be harvested and
summarized. [Worker](https://github.com/andrewraim/worker) is a lightweight
tool that seeks to help automate this process. This post gives a brief
description; see the webpage for details.

Suppose our study consists of a parameter that is varied in some way, and we
have determined that there are 999 levels of interest. Suppose that we have set
up 999 corresponding directories, named as follows.

```{r, eval = FALSE}
$ ls -1
run001
run002
...
run999
```

In each folder, suppose there is a script `launch.R` that runs the
corresponding level of the simulation. An output file `output.csv` is placed in
the directory upon successful completion of `launch.R`. Running the `launch.R`
scripts manually would be extremely tedious, but we could write a loop in our
preferred shell scripting language to automate this.

The worker takes this concept slightly further; it loops through jobs and runs
them sequentially, but multiple workers can be run simultaneously on the same
study to achieve "embarrassingly parallel" parallel computing. Workers make use
of use lock files to avoid stepping on each other's toes. This paradigm can
be useful in shared computing environments, where users may be asked to limit
the number of CPUs requested at any given time. We can start, say, ten workers
rather than having to continually monitor a scheduler's queue and submit more
jobs.

Here is an example of the syntax currently used to invoke each worker in this
example.

```{r, eval = FALSE}
$ worker.sh -p 'run???' -c 'Rscript launch.R'
```

After all of the runs are completed, we will likely want to load each of the
`output.csv` files and construct tables and plots to summarize the results. This
[post](2023-02-11-latex-tables-from-R.qmd) may be helpful if you are looking to
construct Latex tables from a computational study.
