---
title: "Sending Mail from R"
date: 2023-05-01
categories: programming
---

Here is a quick note on sending mail from R. For example, this could be useful
in a long running job to alert yourself when it completes or halts because of
an error. The [sendmailR](https://cran.r-project.org/package=sendmailR) package
provides a general interface to mail.

Here is an example where we use sendmailR to send mail to the local mailbox
of our user (araim) on the host machine. This assumes we have a sendmail
server set up to relay local mail and that araim has a mailbox.

The following call in R sends a message.

```{r}
#| eval: false
result = sendmail(
	from = "sendmailR",
	to = "araim",
	subject = "Hello from R",
	msg = "This is an email from R using the sendmailR package.",
	control = list(smtpServer = "localhost")
)
```

Here is how the message appears when viewed with `mutt`.

```{verbatim}
i:Exit  -:PrevPg  <Space>:NextPg  v:View Attachm.  d:Del  r:Reply  j:Next  ?:Help
Date: Tue, 02 May 2023 09:24:31 -0000
From: sendmailR@localhost
To: araim@localhost
Subject: Hello from R

[-- Attachment #1 --]
[-- Type: text/plain; charset=us-ascii, Encoding: 7bit, Size: 0.1K --]

This is an email from R using the sendmailR package.

-N +- 1155/1155: sendmailR@localhost   Hello from R                      -- (all)
```


Here is an example of sending mail on an exception using `withCallingHandlers`.
Here we use `sys.calls` to give some context about the call that led to the
error.

```{r}
#| eval: false
f = function(x) { stop("An exception!") }
g = function(x) { f(x) }

withCallingHandlers({
	g(1:10)
}, error = function(e) {
	result = sendmail(
		from = "sendmailR",
		to = "araim",
		subject = "Exception from R",
		msg = sprintf("%d: %s", seq_along(sys.calls()), sys.calls()),
		control = list(smtpServer = "localhost")
	)
})
```

And here is how the message appears when viewed with `mutt`.

```{verbatim}
i:Exit  -:PrevPg  <Space>:NextPg  v:View Attachm.  d:Del  r:Reply  j:Next  ?:Help
Date: Tue, 02 May 2023 10:06:56 -0000
From: sendmailR@localhost
To: araim@localhost
Subject: Exception from R

[-- Attachment #1 --]
[-- Type: text/plain; charset=us-ascii, Encoding: 7bit, Size: 0.6K --]

1: withCallingHandlers({
   g(1:10)
}, error = function(e) {
   result = sendmail(from = "sendmailR", to = "araim", subject = "Exception from
+R", msg = sprintf("%d: %s", seq_along(sys.calls()), sys.calls()), control =
+list(smtpServer = "localhost"))
})
2: g(1:10)
3: f(x)
4: stop("An exception!")
5: .handleSimpleError(function (e) {
   result = sendmail(from = "sendmailR", to = "araim", subject = "Exception from
+R", msg = sprintf("%d: %s", seq_along(sys.calls()), sys.calls()), control =
+list(smtpServer = "localhost"))
}, "An exception!", base::quote(f(x)))
6: h(simpleError(msg, call))

-N +- 1162/1162: sendmailR@localhost   Exception from R                  -- (all)
```
